---
title: "Dual Function Recovery: estimation diagnostics"
subtitle: "Bayesian multistate life table, HRS 1998-2022"
format:
  html:
    toc: true
    embed-resources: true
    code-fold: true
    fig-width: 7
    fig-height: 4
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
library(tidyverse); library(here); library(posterior); library(knitr)
source(here("R/functions/constants.R"))
sexes <- c("men", "women"); priors <- c("2.5", "5", "10")
load_fit <- function(sex, psd) readRDS(here("output/models", sprintf("draws_%s_psd%s.rds", sex, psd)))
fits <- map(set_names(sexes), \(s) map(set_names(priors), \(p) load_fit(s, p)))
theme_set(theme_minimal(base_size = 10))
```

This report summarises data construction, sampler convergence, prior sensitivity and life-table checks.
It reads saved posterior draws and tables only; nothing is re-estimated. The main prior is N(0, 5^2).

# 1. Data and sample

```{r sample}
read_csv(here("output/tables/03_sample_flow.csv"), show_col_types = FALSE) |> kable(caption = "Sample flow (reasons for exclusion)")
iv <- readRDS(here("data/derived/intervals.rds"))
tibble(sex = c("Men", "Women", "Total"),
       intervals = c(sum(iv$female == 0), sum(iv$female == 1), nrow(iv)),
       persons = c(n_distinct(iv$hhidpn[iv$female == 0]), n_distinct(iv$hhidpn[iv$female == 1]), n_distinct(iv$hhidpn))) |>
  kable(format.args = list(big.mark = ","), caption = "Analytic sample")
```

```{r transitions}
tc <- iv |> count(start_state, end_state) |> pivot_wider(names_from = end_state, values_from = n, names_prefix = "to ")
kable(tc, format.args = list(big.mark = ","), caption = "Transition counts, men and women combined (to 5 = dead)")
```

```{r bywave}
read_csv(here("output/tables/02_state_by_wave.csv"), show_col_types = FALSE) |>
  ggplot(aes(year, pct, colour = state)) + geom_line() + geom_point(size = 1) +
  labs(x = NULL, y = "% of respondents", colour = NULL, title = "State distribution by wave")
```

# 2. Sampler

Independence Metropolis-Hastings with a t(30) proposal at the posterior mode; 4 chains x 100,000 proposals, burn-in 10,000, thin 10 (36,000 draws per sex).

```{r conv}
conv <- map_dfr(sexes, \(s) map_dfr(priors, \(p) {
  f <- fits[[s]][[p]]; d <- f$diagnostics
  tibble(sex = s, prior_sd = p, N = f$N, acceptance = mean(f$accept), max_rhat = max(d$rhat),
         ess_bulk_median = median(d$ess_bulk), ess_bulk_min = min(d$ess_bulk), ess_tail_min = min(d$ess_tail),
         run_min = as.numeric(f$run_time) / 60)
}))
kable(conv, digits = c(0, 0, 0, 3, 3, 0, 0, 0, 0), format.args = list(big.mark = ","),
      caption = "Convergence by run (190 parameters each)")
```

```{r weakest}
f <- fits$men[["5"]]
bind_rows(map(sexes, \(s) fits[[s]][["5"]]$diagnostics |> mutate(sex = s))) |>
  group_by(sex) |> slice_min(ess_tail, n = 5) |> select(sex, variable, rhat, ess_bulk, ess_tail) |>
  kable(digits = 3, format.args = list(big.mark = ","), caption = "Five parameters with the lowest tail ESS (prior sd 5)")
```

Outcome codes are `(start - 1) * 5 + end`: t2 = 1→2, t5 = 1→5, t8 = 2→3, t12 = 3→2, t16 = 4→1, t18 = 4→3, t20 = 4→5.

```{r trace, fig.height = 6}
chain_df <- function(sex, par) {
  f <- fits[[sex]][["5"]]; ix <- strsplit(par, ":")[[1]]
  x <- f$draws[, , ix[1], ix[2]]
  as_tibble(x, .name_repair = ~ paste0("chain", 1:4)) |> mutate(iter = row_number()) |>
    pivot_longer(-iter, names_to = "chain", values_to = "value") |> mutate(sex = sex, par = par)
}
worst <- map_chr(sexes, \(s) fits[[s]][["5"]]$diagnostics |> slice_min(ess_tail, n = 1) |> pull(variable))
tr <- bind_rows(map2(sexes, worst, chain_df), map(sexes, \(s) chain_df(s, "age:t5")))
ggplot(tr, aes(iter, value, colour = chain)) + geom_line(linewidth = 0.2, alpha = 0.7) +
  facet_grid(par ~ sex, scales = "free_y") + labs(x = "Retained draw (per chain)", y = NULL, title = "Trace plots: weakest-mixing parameter and age:t5 per sex")
```

```{r dens}
ggplot(tr, aes(value, colour = chain)) + geom_density() + facet_grid(par ~ sex, scales = "free") +
  labs(x = NULL, title = "Marginal densities by chain")
```

```{r rhat}
bind_rows(map(sexes, \(s) fits[[s]][["5"]]$diagnostics |> mutate(sex = s))) |>
  ggplot(aes(ess_bulk)) + geom_histogram(bins = 30) + facet_wrap(~ sex, scales = "free_x") + scale_x_log10() +
  labs(x = "Bulk ESS (log scale)", y = "Parameters", title = "Distribution of bulk ESS, prior sd 5")
```

# 3. Prior sensitivity

```{r prior}
ps <- map_dfr(sexes, \(s) read_csv(here("output/tables", sprintf("prior_sensitivity_%s.csv", s)), show_col_types = FALSE) |> mutate(sex = s))
ps |> group_by(sex) |>
  summarise(`sd 2.5 median |diff|` = median(abs(d25)), `sd 2.5 max |diff|` = max(abs(d25)),
            `sd 10 median |diff|` = median(abs(d10)), `sd 10 max |diff|` = max(abs(d10))) |>
  kable(digits = 3, caption = "Posterior-mean differences vs sd 5, in sd-5 posterior SDs")
ps |> mutate(mx = pmax(abs(d25), abs(d10))) |> group_by(sex) |> slice_max(mx, n = 5) |>
  select(sex, par, mean_2.5, mean_5, mean_10, sd_5, d25, d10) |> kable(digits = 2, caption = "Largest shifts")
```

```{r lt-prior}
lt <- map_dfr(sexes, \(s) map_dfr(priors, \(p) read_csv(here("output/tables", sprintf("lifetable_summary_%s_psd%s.csv", s, p)), show_col_types = FALSE)))
lt |> filter(quantity == "total_expectancy", radix == "population", age %in% c(50, 70, 90)) |>
  select(sex, age, prior_sd, mean, lo95, hi95) |>
  kable(digits = 2, caption = "Total life expectancy (population radix) by prior")
```

# 4. Life-table checks

```{r lt-checks}
m5 <- lt |> filter(prior_sd == 5)
m5 |> filter(quantity == "pct_remaining", radix == "population", state != "Total") |>
  group_by(sex, age) |> summarise(s = sum(mean), .groups = "drop") |>
  summarise(`min sum` = min(s), `max sum` = max(s)) |> kable(digits = 6, caption = "Percent of remaining life sums to 100 at every age")
m5 |> filter(quantity == "total_expectancy", radix == "population") |> group_by(sex) |>
  summarise(`monotone decreasing in age` = all(diff(mean[order(age)]) < 0)) |> kable()
m5 |> filter(quantity == "total_expectancy", age == 50) |> select(sex, radix, mean, lo84, hi84) |>
  kable(digits = 2, caption = "Total life expectancy at 50 by radix")
```

# 5. Notes and caveats

- Sampler: Polya-Gamma Gibbs mixed poorly on the full data (pilot: max R-hat 1.58, min ESS 7), so the independence sampler is the main estimator; see `docs/methods-notes.md` sections 8 and 9.
- State 3 → 1 recovery (about 30% per two years) is probably Langa-Weir classification noise near the dementia cut.
- Life tables are cohort tables at the sample-mean birth cohort, with race/ethnicity and education at sample proportions; unweighted.
- Intervals use the 2-year assumption for all intervals. The CIND + dementia and interval-length sensitivity analyses are not yet run.

```{r session}
sessionInfo()
```
